---
title: "Analysing Longitudinal Data"
subtitle: "Fixed Occasions"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warnings = FALSE, comment = NULL, message = FALSE)
```

## Conceptual Issues

First, I list some issues to think about with longitudinal modeling.
With most longitudinal data, we have time points or occasions nested within individuals. 
So, unlike previous data we look at, individuals are now level 2 instead of level 1.

#### Fixed and Varying Occasions

Time can be measured at fixed intervals, in which every individual is measured at the same intervals and or times. 
Alternatively, individuals can be measured at varying times. In this post we will look at longitudinal data with fixed occasions..

### Advantages of Multilevel Longitudinal Modeling

Recall from the text that there are advantages to collecting longitudinal data over cross-sectional data.
These include:

1. greater power with additional time points
2. the ability to look at *how* the outcome changes (patterns of change) and not just *whether* they change.

There are a number of advantages to using multilevel models to model longitudinal data. 
These include:

1. The ability to model growth curves that are different for each individual
2. The number and spacing of occasions can vary between individuals
3. The covariances between occasions can be explicitly modeled.
4. Repeated measures ANOVA is a special (and more restrictive) case of MLM (suggesting that we should start with the less restrictive model: MLM)
5. I is easy to add higher levels to model
6. It is straight-forward to add time-varying or time-constant explanatory variables to the model.

But, the measures must be calibrated across time points. 

## Example with Fixed Occasions

For this example, we are using data from chapter 5 of Hox, Moerbeek, and van de Schoot (2018). 
The data contain the college GPA of 200 college students collected over six occasions. 
A time varying covariate, the job status of each student, was also collected at the six occasions. 
Two time invariant variables, high school GPA and the students gender were also collected. There is another outcome variable named `admitted`, that is dichotomous and indicates whether the student was admitted to the university of their choice.
We will not use this last variable in this post.

The purpose of this example study is to study the impact of having a job on college GPA. 
But, we will want to control for gender and high school GPA in our analyses.

First, we need to load the correct data.
There are at least two ways we can structure 
Load all packages to be used. Note that the data for this project are in SPSS format, so you my need to install the `haven` package which allow for importing SPSS, Stata, and SAS data files into R.
So, you may need to run `install.packages('haven')` before running the code chunk below.
```{r}
# install the haven package which allows importing SPSS data, among other 
# formats (Stata, SAS)
library(haven) 
library(lme4)
library(lmerTest)
library(ggplot2)
library(texreg)
library(performance)
library(interactions)
```

##  Import Data

Import the `gpa2long.sav` file from the data directory of the **longidtudinal_pgaCurran** project.
```{r, eval=FALSE}
gpa2long <- read_sav("data/gpa2long.sav")
```

```{r, echo=FALSE}
gpa2long <- read_sav("projects/longitudinalMLM_fixedTiime/data/gpa2long.sav")
```

Note that the variables in this data have preserved the SPSS labeling, which can be used to determine the factor labels. 
To see this use the `str()` function on the data frame.

## Clean Data

The data is in good shape, all we need to do is convert the categorical variables to factors.
Here, I leave occasion as numeric, to reflect the assumption that this variable measure evenly spaced time points, that start with zero at the first time point. 
```{r}
gpa2long <- within(gpa2long, {
  student <- factor(student)
  gender <- factor(sex, labels = c("male", "female"))
  admitted <- factor(admitted, labels = c("no", "yes"))
})
```

## Explore Data

This first plot is a spaghetti plot, which just plots the outcome at each occasion with the points for each individual connected with a line. Its hard to see what is going on with each individual, but we can see a general trend in increases in GPA across occasion.

```{r, warning=FALSE}
ggplot(gpa2long, aes(y = gpa, x = occas, group = student)) + 
  geom_line()
```

Another way to look at the general trend is to get a smoothed line summarizing the trend across individuals. 

```{r}
ggplot(gpa2long, aes(y = gpa, x = occas)) +
  geom_point() + geom_smooth(method = "loess")
```

The loess smoothing algorithm allows for a local average curve to be fit to the data, which would allow for a curvy line. 
The fact that the line looks very straight suggest a linear relation between gpa across occasions.
This plot also demonstrate the fixed nature of the occasions, as they line up in descrete columns.

We can also look at the linear relation across occasions for each individual.

```{r}
ggplot(gpa2long, aes(y = gpa, x = occas, group = student)) + 
  geom_smooth(method = "lm", se = FALSE)
```

Here we see that descriptively our sample demonstrates differences in the estimated slopes across individuals. 
There even seem to be some that are negative.

We can look at the mean GPA across time in a few ways:

```{r}
boxplot(gpa ~ occas, gpa2long)
```

This plot actually shows the medians and other quartiles, but we see the upward trend in GPA.

Next, I calculate the mean GPA for each occasion, and print and plot them.

```{r}
gpameans_occas <- aggregate(gpa ~ occas, gpa2long, mean)
gpameans_occas
plot(gpa ~ occas, gpameans_occas)
```




### **Exercise**
1. Create some more plots exploring the relation of key variables with other variables in the data.
2. How would you summarize what you have learned about the data to this point?
3. Obtain some summary statistics of the data, now that you have a sense of data as a whole. Based on what you learned from the graphic exploration, are these statistics reasonable summaries of this data?

## Models

Now we are ready to run a series of model of increasing complexity.

```{r}
mod0 <- lmer(gpa ~ 1 + (1 | student), gpa2long, REML = FALSE)
modtime <- lmer(gpa ~ occas + (1 | student), gpa2long, REML = FALSE)
modtjob <- lmer(gpa ~ occas + job + (1 | student), gpa2long, REML = FALSE)
modfull <- lmer(gpa ~ occas + job + highgpa + gender + (1 | student), gpa2long,
                REML = FALSE)
summary(modfull)
```

```{r, results = 'asis'}

htmlreg(list(mod0, modtime, modtjob, modfull), digits = 3, 
        custom.model.names = c("$M_1$: null model", "$M_2$:+occas.",
                               "$M_3$:+jobsat", "$M_4$:+gpa, gender"))
```


#### Testing Nonlinear trends in GPA growth

```{r}
modtimepoly <- lmer(gpa ~ poly(occas, 2) + (1 | student), gpa2long, REML = FALSE)
summary(modtimepoly)
```

## Adding Random Slopes

```{r}
modfullrs <- lmer(gpa ~ occas + job + I(highgpa - mean(highgpa)) + gender + (occas |student), 
                  gpa2long, REML = FALSE)
summary(modfullrs)
```

```{r, results = 'asis'}
htmlreg(list(modtime, modfullrs), 
        custom.model.names = c("$M_5$:+occas rand", 
                               "$M_6$:+cross-level int."))
```

### Test the addition of the random slope for occasions.

```{r}
ranova(modfullrs)
anova(modfull, modfullrs) # equivalent
```

## Cross-level Interaction occassion*gender

```{r}
modfullrsx <- lmer(gpa ~ occas + job + I(highgpa - mean(highgpa)) + gender + 
                                          gender:occas + 
                     (occas |student), 
                  gpa2long, REML = FALSE)
```

```{r, results='asis'}
htmlreg(list(modfull, modfullrs, modfullrsx), digits = 3)
```

### Get correlation between variance components 
```{r}

VarCorr(modfullrsx)
```

### Figure 5.4, p. 82.

```{r}
interact_plot(modfullrsx, occas, gender)
```

```{r}
sim_slopes(modfullrsx, occas, gender)
```

## What about smaller sample size (at level 2)? 

```{r}

# create a smaller data set by sampling individuals from the full data:
n <- 12 # level 2 sample size. You can change this to play around with n.
# Index of sampled students
set.seed(20200702) # set the random seed for reproducibility
sampidx <- as.numeric(sample(unique(gpa2long$student), replace = FALSE, size = n)) 

# Create smaller sample data
gpa2long12 <- gpa2long[gpa2long$student %in% sampidx, ]
```


## References

Hox, J.J., Moerbeek, M. and Van de Schoot, R. (2018), Multilevel Analysis: Techniques and applications, Sage, New York, NY.