---
title: 'Longitudinal Modeling' 
subtitle: 'Alternative Covariance Structures'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = NULL)
```

```{r}
# packages used -----------------------------------------------------------
library(tidyverse)  
library(car)
library(psych)
library(nlme)
```


Alternative Covariance Structures form a continuum from the most constrained to the least constrained varaince/covariance patter imposed by the model. 


## Data Example

```{r}
mood <- haven::read_dta("data/STATA_Chapter4.dta")

mood <- transform(mood,
                  studyday = factor(studyday),
                  personid = factor(personid)
)

```

### Create wide dataframe

```{r}
moodwide <- reshape(mood, direction = "wide", v.names = "posmood",
                     idvar = "personid", timevar = "studyday")
```

## Explore Mood Data

```{r}
round(cov(moodwide[ ,-1], use = "complete.obs"), 3)
```

### Plot Individual Trajectories

```{r}
ggplot(mood, aes(studyday, posmood, group = personid, color = personid)) + 
  geom_line() +  # geom_point() +
  stat_summary(fun = mean, geom = "line", lwd = 2, color = "red",
               aes(group = 1)) + 
  theme(legend.position = "none")
```

## Unstructured 

```{r}
unstruct <- gls(posmood ~ studyday, mood, 
                correlation=corSymm(form = ~ 1 |personid),  
                weights=varIdent(form = ~ 1|studyday),method = "REML")
summary(unstruct)
```

```{r}
getVarCov(unstruct)
```

## Compound Symmetry

```{r}
comsym <- gls(posmood ~ studyday, mood, 
              correlation=corCompSymm(form = ~ 1 |personid),
              method="REML")
summary(comsym)
```

```{r}
getVarCov(comsym)
comsym$modelStruct$corStruct

# Visualize the full correlation matrix
cc <- corMatrix(comsym$modelStruct$corStruct)[[5]]
round(cc, 2)
```

## Compound Symmetry, Heterogeneous 
```{r}
# CS Heterogeneous --------------------------------------------------------

hetercom <- gls(posmood ~ studyday, mood, 
                correlation=corCompSymm(form = ~ 1 |personid),
                weights=varIdent(form = ~1|studyday), 
                method="REML")
getVarCov(hetercom)
```


## Autoregressive
```{r}
# Autoregressive ----------------------------------------------------------

auto1 <- gls(posmood ~ studyday, mood, 
             correlation=corAR1(form = ~ 1 |personid),
             method="REML")
cc <- corMatrix(auto1$modelStruct$corStruct)[[5]]
print(cc)

cc * auto1$sigma^2
```

## Autoregressive, Heterogeneous
```{r}
# Autoregressive Heterogeneous --------------------------------------------

hauto1 <- gls(posmood ~ studyday, mood, 
              correlation=corAR1(form = ~ 1 |personid), 
              weights=varIdent(form = ~1|studyday), method="REML")
getVarCov(hauto1)
```

## Toeplitz
```{r}
# Toeplitz ----------------------------------------------------------------

toep <- gls(posmood ~ studyday, mood, 
            correlation=corARMA(form = ~ 1 |personid, p=3, q=0), 
            method="REML")
cc <- corMatrix(toep$modelStruct$corStruct)[[5]]
print(cc)
cc * toep$sigma^2
```

## Compare Models
```{r}

anova(comsym,hetercom,auto1,hauto1,toep, unstruct)
```
